---
title: "metaXpress Full Workflow: GEO to Report"
author: "Md. Jubayer Hossain"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{metaXpress Full Workflow: GEO to Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  eval     = FALSE,  # All chunks are illustrative; network calls not run in CI
  purl     = FALSE   # Exclude from R CMD check vignette code extraction
)
```

# Overview

`metaXpress` is an end-to-end R package designed for multi-study transcriptomic meta-analysis. It provides a standardized pipeline covering data ingestion from NCBI GEO and SRA, 10-point quality control scoring, cross-platform harmonization, per-study differential expression, six statistical meta-analysis models, missing gene handling, pathway enrichment, interactive visualization, and reproducible reporting.

For full documentation and online tutorials, visit the official website at [https://mdjubayerhossain.com/metaXpress/](https://mdjubayerhossain.com/metaXpress/).

# Installation from GitHub

`metaXpress` is currently available on GitHub and targets future submission to Bioconductor.

```{r install-github, eval=FALSE}
# 1. Ensure BiocManager is present for Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# 2. Install remotes
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

# 3. Install metaXpress
remotes::install_github("hossainlab/metaXpress", dependencies = TRUE)
```

Load the package:

```{r load-package, purl=FALSE}
library(metaXpress)
```

---

# 1. Data Ingestion & Quality Control

## 1a. Fetch Public Data from GEO

You can ingest multiple public RNA-seq datasets directly using their NCBI GEO accessions:

```{r fetch-geo, purl=FALSE}
# Fetch three bulk RNA-seq cohorts from GEO
accessions <- c("GSE130688", "GSE136569", "GSE171485")
studies    <- mx_fetch_geo(accessions, count_type = "raw", cache_dir = "geo_cache/")
```

Alternatively, load local count matrices and metadata with `mx_load_local()`.

## 1b. Inspect 10-Point QC Scoring

`mx_qc_study()` benchmarks each study against 10 criteria (sample size, depth, alignment rate, duplicate rate, case/control completeness, integer counts, etc.):

```{r qc-summary, purl=FALSE}
qc_scores <- vapply(studies, function(s) s@qc_score, numeric(1))
print(qc_scores)

# Inspect detailed metrics for study 1
attr(studies[[1]], "qc_details")
```

## 1c. Filter Low-Quality Cohorts

```{r filter-studies, purl=FALSE}
# Filter out cohorts failing the QC threshold (minimum score: 7/10)
studies <- mx_filter_studies(studies, qc_threshold = 7)
```

---

# 2. Cross-Study Harmonization

## 2a. Reannotate Gene Identifiers

Convert diverse gene identifiers (Ensembl, Entrez, RefSeq) to a uniform namespace:

```{r reannotate, purl=FALSE}
studies <- mx_reannotate(studies, org = "Homo sapiens", target_id = "SYMBOL")
```

## 2b. Correct Library Type Bias

If cohorts mix poly(A)-selected and rRNA-depleted protocols, adjust for library type biases:

```{r library-correction, purl=FALSE}
studies <- mx_correct_library_type(studies)
```

## 2c. Remove Batch Effects

```{r batch-correction, purl=FALSE}
studies <- mx_remove_batch(studies, method = "ComBat-seq")
```

## 2d. Align Common Gene Space

```{r align-genes, purl=FALSE}
studies <- mx_align_genes(studies)
```

---

# 3. Per-Study Differential Expression

Run independent statistical tests on each cohort using DESeq2, edgeR, or limma-voom:

```{r de-analysis, purl=FALSE}
studies <- mx_de_all(
  studies,
  method  = "DESeq2",
  formula = ~ condition,
  BPPARAM = BiocParallel::MulticoreParam(4)
)
```

Inspect significant gene counts:

```{r de-summary, purl=FALSE}
mx_de_summary(studies, padj_threshold = 0.05, lfc_threshold = 1.0)
```

---

# 4. Handle Missing Genes

When cohorts have varying gene detection rates:

```{r missing-genes, purl=FALSE}
de_results <- lapply(studies, function(s) s@de_result)

# Inspect gene coverage across cohorts
cov_mat <- mx_missing_summary(de_results)

# Retain genes present in at least 2 cohorts or apply k-NN imputation
de_results <- mx_filter_coverage(de_results, min_studies = 2)
# de_results <- mx_impute(de_results, method = "knn")
```

---

# 5. Statistical Meta-Analysis

## 5a. Model Selection Guide

- **2 cohorts:** `"fisher"` (Fisher's combined probability) or `"stouffer"` (Z-score weighting).
- **3+ cohorts, low heterogeneity ($I^2 < 25\%$):** `"fixed_effects"` (Inverse-variance).
- **3+ cohorts, moderate-to-high heterogeneity ($I^2 \ge 25\%$):** `"random_effects"` (DerSimonian-Laird, recommended default).
- **Mixed heterogeneity:** `"awmeta"` (Adaptive weighting).

```{r meta-analysis, purl=FALSE}
meta_result <- mx_meta(de_results, method = "random_effects", min_studies = 2)
meta_result
```

## 5b. Quantify Between-Study Heterogeneity

Assess Cochran's $Q$, Higgins $I^2$, and $\tau^2$ for every gene:

```{r heterogeneity, purl=FALSE}
het <- mx_heterogeneity(de_results)
summary(het$I_sq)
```

## 5c. Leave-One-Out Sensitivity Analysis

Ensure discoveries are not driven by an individual outlier cohort:

```{r sensitivity, purl=FALSE}
loo <- mx_sensitivity(de_results, method = "random_effects")
vapply(loo, function(r) sum(r@meta_table$meta_padj <= 0.05, na.rm = TRUE), integer(1))
```

---

# 6. Pathway Meta-Analysis

Perform cross-study Over-Representation Analysis (ORA) or Gene Set Enrichment Analysis (GSEA) on Hallmark, KEGG, or GO sets:

```{r pathway, purl=FALSE}
meta_result <- mx_pathway_meta(meta_result, db = "Hallmarks", method = "ORA")
head(meta_result@pathway_result[order(meta_result@pathway_result$padj), ], 10)
```

---

# 7. Publication Visualizations

## Volcano Plot

```{r volcano, fig.width=8, fig.height=6, purl=FALSE}
mx_volcano(meta_result, padj_threshold = 0.05, lfc_threshold = 1.0, label_top = 15)
```

## Forest Plot

```{r forest-top-gene, fig.width=7, fig.height=5, purl=FALSE}
top_gene <- meta_result@meta_table$gene_id[which.min(meta_result@meta_table$meta_padj)]
mx_forest(top_gene, de_results, meta_result)
```

## Heterogeneity Plot

```{r i2-distribution, fig.width=6, fig.height=4, purl=FALSE}
mx_heterogeneity_plot(meta_result)
```

---

# 8. Interactive Explorer & Report Generation

## Launch Interactive Shiny App

Explore results dynamically in your browser:

```{r run-app, eval=FALSE}
metaXpress::mx_run_app()
```

## Generate Reproducible Reports & Export

```{r report, purl=FALSE}
mx_report(meta_result, studies, de_results, format = "html", output_dir = "results/")
mx_export(meta_result, format = "csv", output_dir = "results/")
```

---

# Session Information

```{r session-info, eval=FALSE, purl=FALSE}
mx_session_info()$session_info
```
