---
title: "metaXpress Full Workflow: GEO to Report"
author: "DeepBio Academy"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{metaXpress Full Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  eval     = FALSE   # Network calls disabled in vignette build
)
```

# Overview

This vignette demonstrates the complete metaXpress workflow, from fetching
studies from GEO through pathway enrichment and report generation.
All `mx_fetch_*()` calls require internet access and are shown with
`eval=FALSE`; substitute `data(metaXpress_example)` for local testing.

```{r load-package}
library(metaXpress)
```

# 1. Data Ingestion

## 1a. Fetch from GEO

```{r fetch-geo}
# Fetch three Alzheimer's disease RNA-seq studies
accessions <- c("GSE53697", "GSE95587", "GSE118553")
studies    <- mx_fetch_geo(accessions, count_type = "raw",
                            cache_dir = "geo_cache/")
```

## 1b. Inspect QC Results

```{r qc-summary}
qc_scores <- vapply(studies, function(s) s@qc_score, numeric(1))
print(qc_scores)

# View detailed QC breakdown for the first study
attr(studies[[1]], "qc_details")
```

## 1c. Filter Low-Quality Studies

```{r filter-studies}
studies <- mx_filter_studies(studies, qc_threshold = 7)
```

# 2. Harmonization

## 2a. Reannotate Gene IDs

```{r reannotate}
studies <- mx_reannotate(studies, org = "Homo sapiens",
                          target_id = "SYMBOL")
```

## 2b. Correct Library Type Bias

If studies mix polyA-selected and rRNA-depleted libraries:

```{r library-correction}
studies <- mx_correct_library_type(studies)
```

## 2c. Remove Batch Effects

```{r batch-correction}
studies <- mx_remove_batch(studies, method = "ComBat-seq")
```

## 2d. Align to Common Genes

```{r align-genes}
studies <- mx_align_genes(studies)
```

# 3. Per-Study Differential Expression

```{r de-analysis}
studies <- mx_de_all(studies, method = "DESeq2",
                      formula = ~ condition,
                      BPPARAM = BiocParallel::MulticoreParam(4))
```

```{r de-summary}
mx_de_summary(studies, padj_threshold = 0.05, lfc_threshold = 1)
```

# 4. Handle Missing Genes

```{r missing-genes}
de_results <- lapply(studies, function(s) s@de_result)

# Check coverage across studies
cov_mat <- mx_missing_summary(de_results)
hist(attr(cov_mat, "coverage_pct"),
     main = "Gene coverage across studies",
     xlab = "% studies with gene detected")

# Keep genes in at least 2 of 3 studies
de_results <- mx_filter_coverage(de_results, min_studies = 2)
```

# 5. Meta-Analysis

## 5a. Choose the Right Method

Use the decision guide:
- 2 studies → `"fisher"` or `"stouffer"`
- 3+ studies, low heterogeneity → `"fixed_effects"`
- 3+ studies, high heterogeneity → `"random_effects"` (default)
- Mixed heterogeneity → `"awmeta"`

```{r meta-analysis}
meta_result <- mx_meta(de_results, method = "random_effects",
                        min_studies = 2)
meta_result
```

## 5b. Heterogeneity Assessment

```{r heterogeneity}
het <- mx_heterogeneity(de_results)
summary(het$I_sq)

# Genes with high heterogeneity
high_het <- het[!is.na(het$I_sq) & het$I_sq > 75, ]
nrow(high_het)
```

## 5c. Sensitivity Analysis

```{r sensitivity}
loo <- mx_sensitivity(de_results, method = "random_effects")
# Compare significant gene counts across LOO runs
vapply(loo, function(r) sum(r@meta_table$meta_padj <= 0.05, na.rm = TRUE),
        integer(1))
```

# 6. Pathway Meta-Analysis

```{r pathway}
meta_result <- mx_pathway_meta(meta_result, db = "Hallmarks",
                                method = "ORA")
head(meta_result@pathway_result[
  order(meta_result@pathway_result$padj), ], 10)
```

# 7. Visualization

```{r volcano, fig.width=8, fig.height=6}
mx_volcano(meta_result, padj_threshold = 0.05, lfc_threshold = 1,
            label_top = 15)
```

```{r forest-top-gene, fig.width=7, fig.height=5}
top_gene <- meta_result@meta_table$gene_id[
  which.min(meta_result@meta_table$meta_padj)]
mx_forest(top_gene, de_results, meta_result)
```

```{r heatmap, fig.width=9, fig.height=8}
mx_heatmap(meta_result, studies, top_n = 30)
```

```{r i2-distribution, fig.width=6, fig.height=4}
mx_heterogeneity_plot(meta_result)
```

# 8. Report & Export

```{r report}
mx_report(meta_result, studies, de_results,
           format     = "html",
           output_dir = "results/")
```

```{r export}
mx_export(meta_result, format = "excel", output_dir = "results/")
mx_export(meta_result, format = "csv",   output_dir = "results/")
```

# Session Information

```{r session-info, eval=TRUE}
mx_session_info()$session_info
```
