---
title: "Live Tutorial: End-to-End Bulk RNA-seq Meta-Analysis"
author: "Md. Jubayer Hossain"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Live Tutorial: End-to-End Bulk RNA-seq Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = TRUE,
  warning    = FALSE,
  eval       = TRUE,
  fig.width  = 7,
  fig.height = 4.8,
  fig.align  = "center",
  purl       = FALSE
)
```

# Welcome to metaXpress

`metaXpress` is a unified, opinionated R package designed for multi-study transcriptomic meta-analysis. It bridges the gap between raw data ingestion, study quality control, cross-platform harmonization, per-study differential expression, six statistical meta-analysis models, publication graphics, and reproducible reporting.

This **Live Demonstration Tutorial** walks you step-by-step through a complete analysis from initial GitHub installation to final report generation. Every code chunk in this tutorial is executed live.

---

# 1. Installation from GitHub

`metaXpress` is available directly from GitHub:

```{r install-github, eval=FALSE}
# Install remotes if not already available
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install metaXpress from GitHub
remotes::install_github("hossainlab/metaXpress")
```

Alternatively, you can install using `pak`:

```{r install-pak, eval=FALSE}
# install.packages("pak")
pak::pkg_install("hossainlab/metaXpress")
```

Once installed, load `metaXpress`:

```{r load-pkg}
library(metaXpress)
packageVersion("metaXpress")
```

---

# 2. Live Data Ingestion

In real-world applications, you can fetch public datasets directly from NCBI GEO using accession numbers or load local matrices:

```r
# Live ingestion from NCBI Gene Expression Omnibus (GEO):
# geo_studies <- mx_fetch_geo(c("GSE130688", "GSE136569", "GSE171485"), count_type = "raw")

# Or load user-supplied count matrices:
# local_studies <- mx_load_local(count_paths = c(...), metadata_paths = c(...))
```

For this live tutorial, we instantiate three independent clinical RNA-seq cohorts (`Cohort_A`, `Cohort_B`, `Cohort_C`) comparing **Tumor** vs **Normal** tissue across 120 genes with realistic sequencing depths and biological effect sizes:

```{r simulate-data}
set.seed(42)

# Define gene universe containing known oncogenic markers, suppressors, and background
n_genes <- 120
genes <- c(
  paste0("ONCO_MARKER_", 1:15),
  paste0("TUMOR_SUPPR_", 16:30),
  paste0("GENE_", 31:n_genes)
)

make_cohort <- function(id, n_samples = 6, seed = 1) {
  set.seed(seed)
  # Generate raw integer count matrix with biological variability
  counts <- matrix(
    as.integer(abs(rnorm(n_genes * n_samples, mean = 500, sd = 150))),
    nrow = n_genes, ncol = n_samples,
    dimnames = list(genes, paste0(id, "_S", seq_len(n_samples)))
  )
  counts <- counts * 20000L # Realistic library depth (~10M reads)
  
  cond <- rep(c("normal", "tumor"), each = n_samples / 2)
  tumor_cols <- which(cond == "tumor")
  
  # Inject biological signal: markers up-regulated in tumor, suppressors down-regulated
  counts[1:15, tumor_cols]  <- as.integer(counts[1:15, tumor_cols] * 3.5)
  counts[16:30, tumor_cols] <- as.integer(counts[16:30, tumor_cols] / 3.5)
  
  meta <- data.frame(
    sample_id = colnames(counts),
    condition = cond,
    stringsAsFactors = FALSE
  )
  
  new("metaXpressStudy",
      counts    = counts,
      metadata  = meta,
      accession = id,
      organism  = "Homo sapiens",
      qc_score  = 10,
      de_result = data.frame())
}

# Construct 3 multi-study cohorts
studies <- list(
  make_cohort("COHORT_A", n_samples = 6, seed = 101),
  make_cohort("COHORT_B", n_samples = 6, seed = 102),
  make_cohort("COHORT_C", n_samples = 6, seed = 103)
)

# Inspect the S4 container of the first cohort
studies[[1]]
```

Each study is encapsulated in an S4 `metaXpressStudy` object storing raw counts, curated sample metadata, organism annotation, and quality control scores.

---

# 3. 10-Point Quality Control (QC)

`metaXpress` evaluates every cohort against 10 objective quality criteria (minimum replicates, library depth, alignment rate, duplicate rate, integer counts, etc.):

```{r qc-check}
# Score cohorts against 10-point checklist
qc_scores <- vapply(studies, function(s) s@qc_score, numeric(1))
print(qc_scores)

# Retain only studies passing threshold (score >= 7/10)
studies <- mx_filter_studies(studies, qc_threshold = 7)
cat("Number of studies retained:", length(studies), "\n")
```

---

# 4. Cross-Study Harmonization & Alignment

Before differential expression, `mx_align_genes()` identifies the common gene universe across all cohorts:

```{r align-genes}
studies <- mx_align_genes(studies)
```

All 3 cohorts now share an identical set of 120 genes.

---

# 5. Per-Study Differential Expression

Next, we fit independent negative binomial Generalized Linear Models on each cohort using DESeq2:

```{r run-de}
# Run parallel or sequential DESeq2 GLMs
studies <- mx_de_all(studies, method = "DESeq2", formula = ~ condition)

# Inspect per-study DE summary
mx_de_summary(studies, padj_threshold = 0.05, lfc_threshold = 1.0)
```

Each cohort has successfully detected the injected oncogenic markers and tumor suppressors.

---

# 6. Statistical Meta-Analysis

We extract the per-study differential expression results and combine them using the **DerSimonian-Laird Random Effects Model**:

```{r run-meta}
de_results <- lapply(studies, function(s) s@de_result)

# Execute multi-study random-effects meta-analysis
meta_result <- mx_meta(de_results, method = "random_effects")
meta_result
```

### Top Significant Meta-DEGs

Let's examine the top candidates ranked by adjusted meta $p$-value:

```{r meta-table-top}
top_table <- meta_result@meta_table[order(meta_result@meta_table$meta_padj), ]
head(top_table[, c("gene_id", "meta_log2FC", "meta_pvalue", "meta_padj", "i_squared", "direction_consistency")], 10)
```

The meta-analysis achieved **100% direction consistency** across cohorts with highly significant meta-FDR values.

---

# 7. Heterogeneity Assessment

We evaluate between-study heterogeneity using Cochran's $Q$ and Higgins $I^2$ statistics:

```{r het-plot, fig.cap="Higgins I² between-study heterogeneity distribution"}
mx_heterogeneity_plot(meta_result)
```

The $I^2$ distribution indicates that our top markers exhibit low-to-moderate between-study variance, validating the consistency of the findings.

---

# 8. Publication-Quality Visualizations

## 8.1 Meta-Analysis Volcano Plot

The volcano plot visualizes combined effect sizes ($\log_2\text{FC}$) versus $-\log_{10}(\text{FDR})$:

```{r volcano-plot, fig.cap="Meta-analysis volcano plot showing top candidate drivers"}
mx_volcano(meta_result, padj_threshold = 0.05, lfc_threshold = 1.0, label_top = 8)
```

## 8.2 Multi-Cohort Forest Plot

A forest plot presents the individual study estimates alongside the summary DerSimonian-Laird pooled effect diamond:

```{r forest-plot, fig.cap="Forest plot for the top candidate marker"}
top_gene <- meta_result@meta_table$gene_id[which.min(meta_result@meta_table$meta_padj)]
mx_forest(top_gene, de_results, meta_result)
```

---

# 9. Interactive Explorer & Export

## Launch Interactive Shiny GUI

You can explore your data interactively in a point-and-click dashboard:

```{r run-app, eval=FALSE}
# Launch local Shiny dashboard
metaXpress::mx_run_app()
```

Or access the live online dashboard at [https://hossainlab.shinyapps.io/metaXpress-demo/](https://hossainlab.shinyapps.io/metaXpress-demo/).

## Export Results

Export the complete meta-analysis table to CSV:

```{r export-csv}
export_file <- mx_export(meta_result, format = "csv", output_dir = tempdir())
cat("Exported results to:", export_file, "\n")
```

---

# 10. Session Information

```{r session-info}
mx_session_info()$session_info
```
