---
title: "Live Tutorial: End-to-End GEO Pan-Cancer Meta-Analysis"
author: "Md. Jubayer Hossain"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Live Tutorial: End-to-End GEO Pan-Cancer Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = TRUE,
  warning    = FALSE,
  eval       = TRUE,
  fig.width  = 7,
  fig.height = 4.8,
  fig.align  = "center",
  purl       = FALSE
)
```

# Welcome to metaXpress

`metaXpress` is a unified, opinionated R package for multi-study transcriptomic meta-analysis. It covers data ingestion, 10-point quality control scoring, cross-platform harmonization, per-study differential expression, six statistical meta-analysis models, publication-grade graphics, and reproducible reporting.

This **Live Demonstration Tutorial** walks you step-by-step through a complete analysis on real-world clinical transcriptomics:
* **Colorectal Adenocarcinoma (CRC):** [GSE130688](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE130688) ($n = 30$ patient samples: 15 normal vs 15 tumor)
* **Pancreatic Ductal Adenocarcinoma (PDAC):** [GSE136569](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE136569) ($n = 10$ patient samples: 5 normal vs 5 tumor)
* **Clear Cell Renal Cell Carcinoma (ccRCC):** [GSE171485](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171485) ($n = 12$ patient samples: 6 normal vs 6 tumor)

Every step and visualization in this tutorial is executed live in real time.

---

# 1. Installation from GitHub

`metaXpress` is installable directly from GitHub:

```{r install-github, eval=FALSE}
# Install remotes if not already available
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install metaXpress directly from GitHub
remotes::install_github("hossainlab/metaXpress")
```

Once installed, load `metaXpress`:

```{r load-pkg}
library(metaXpress)
packageVersion("metaXpress")
```

---

# 2. Ingesting Real Public Cohorts from NCBI GEO

In practice, you can fetch public RNA-seq count matrices and metadata directly from NCBI GEO using their accession numbers:

```{r fetch-geo-syntax, eval=FALSE}
# Live ingestion directly from NCBI Gene Expression Omnibus (GEO):
accessions <- c("GSE130688", "GSE136569", "GSE171485")
studies    <- mx_fetch_geo(accessions, count_type = "raw", cache_dir = "geo_cache/")
```

For this live demonstration, we load the real case study cohort data extracted from these three NCBI GEO series across **52 patient samples**:

```{r load-real-cohorts}
# Load the real case study cohorts
data_path <- system.file("extdata", "case_study_subset.rds", package = "metaXpress")
if (!file.exists(data_path)) {
  data_path <- "inst/extdata/case_study_subset.rds"
}

studies <- readRDS(data_path)

# Verify loaded cohorts
print(names(studies))

# Sample size breakdown per cohort
vapply(studies, function(s) ncol(s@counts), integer(1))
```

Let's inspect the S4 container for the first cohort ([GSE130688](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE130688)):

```{r inspect-s4}
studies[[1]]
```

Each cohort is stored in a clean S4 `metaXpressStudy` object holding raw integer counts, patient phenotype metadata, and quality scores.

---

# 3. 10-Point Study Quality Control (QC)

`metaXpress` evaluates every cohort against 10 objective criteria (minimum sample size, sequencing depth, alignment rate, duplicate rate, integer counts, etc.):

```{r qc-check}
# Inspect 10-point QC checklist scores
qc_scores <- vapply(studies, function(s) s@qc_score, numeric(1))
print(qc_scores)

# Filter cohorts passing QC threshold (score >= 7/10)
studies <- mx_filter_studies(studies, qc_threshold = 7)
cat("Number of cohorts retained:", length(studies), "out of 3\n")
```

All three independent public cohorts scored a perfect 10/10 and were retained.

---

# 4. Cross-Study Harmonization & Common Gene Alignment

Before downstream analysis, `mx_align_genes()` restricts each study matrix to the common gene universe:

```{r align-genes}
studies <- mx_align_genes(studies)
```

---

# 5. Per-Study Differential Expression

We fit negative binomial Generalized Linear Models independently on each study using **DESeq2**:

```{r run-de}
# Fit DESeq2 GLMs across all cohorts
studies <- mx_de_all(studies, method = "DESeq2", formula = ~ condition)

# Inspect DEG detection across cohorts
mx_de_summary(studies, padj_threshold = 0.05, lfc_threshold = 1.0)
```

### The Statistical Power Bottleneck

Notice the stark difference in statistical detection among individual cohorts:
* **GSE130688 ($n=30$):** High statistical power, detecting 35 significant DEGs in this subset.
* **GSE136569 ($n=10$):** Detecting only 2 DEGs after FDR correction.
* **GSE171485 ($n=12$):** Detecting 0 DEGs individually due to severe multiple testing penalties on small sample sizes.

This illustrates why multi-study meta-analysis is essential: combining effect sizes and standard errors across cohorts rescues underpowered discoveries.

---

# 6. Multi-Study Meta-Analysis (Random Effects Model)

We extract the individual cohort differential expression results and pool them using the **DerSimonian-Laird Random Effects Model**:

```{r run-meta}
de_results <- lapply(studies, function(s) s@de_result)

# Execute DerSimonian-Laird random effects meta-analysis
meta_result <- mx_meta(de_results, method = "random_effects")
meta_result
```

Notice how `metaXpress` amplified statistical power: **genes that failed significance in individual small cohorts are successfully recovered** in the pooled meta-analysis.

---

# 7. Biological Annotation & Conserved Cancer Drivers

We map the NCBI Entrez identifiers to canonical gene symbols:

```{r annotate-genes}
# NCBI Entrez to Gene Symbol dictionary for top cancer markers
gene_map <- c(
  "4486"   = "MST1R",
  "4585"   = "MUC4",
  "10643"  = "IGF2BP3",
  "64866"  = "CDCP1",
  "11170"  = "FAM107A",
  "6723"   = "SRM",
  "7498"   = "XDH",
  "623"    = "BDKRB1",
  "284612" = "SYPL2",
  "7274"   = "TTPA"
)

mt <- meta_result@meta_table
mt$symbol <- gene_map[as.character(mt$gene_id)]
mt$symbol[is.na(mt$symbol)] <- mt$gene_id[is.na(mt$symbol)]
meta_result@meta_table <- mt

# Top significant conserved cancer drivers (FDR < 0.05)
key_results <- mt[mt$gene_id %in% names(gene_map), 
                  c("gene_id", "symbol", "meta_log2FC", "meta_pvalue", "meta_padj", "i_squared", "direction_consistency")]
key_results <- key_results[order(key_results$meta_padj), ]
key_results
```

### Canonical Cancer Drivers Recovered:
1. **`MST1R` (RON):** Receptor tyrosine kinase driving cell motility, invasion, and epithelial-to-mesenchymal transition ($\text{meta }\log_2\text{FC} = +1.39, \text{FDR} = 2.39 \times 10^{-5}$).
2. **`MUC4`:** Mucinous glycoprotein promoting tumor proliferation and anti-apoptosis ($\text{meta }\log_2\text{FC} = +2.44, \text{FDR} = 1.51 \times 10^{-6}$).
3. **`IGF2BP3` (IMP3):** Oncofetal RNA-binding protein promoting translation of proliferative transcripts ($\text{meta }\log_2\text{FC} = +2.02, \text{FDR} = 1.03 \times 10^{-5}$).
4. **`CDCP1`:** Transmembrane glycoprotein mediating anoikis resistance ($\text{meta }\log_2\text{FC} = +0.81, \text{FDR} = 1.12 \times 10^{-3}$).
5. **`FAM107A` (DRR1):** Documented tumor suppressor lost during carcinogenesis ($\text{meta }\log_2\text{FC} = -2.48, \text{FDR} = 6.49 \times 10^{-8}$).

All candidate drivers exhibit **100% direction consistency** across independent solid tumors with zero-to-low between-study heterogeneity ($I^2 \le 11\%$).

---

# 8. Between-Study Heterogeneity ($I^2$) Assessment

We examine the Higgins $I^2$ distribution across all evaluated genes:

```{r het-plot, fig.cap="Higgins I² between-study heterogeneity index distribution across cohorts"}
mx_heterogeneity_plot(meta_result)
```

The majority of genes show low between-study heterogeneity ($I^2 < 25\%$), confirming high concordance across the three clinical tumor types.

---

# 9. Publication-Quality Visualizations

## 9.1 Pan-Cancer Meta-Analysis Volcano Plot

```{r volcano-plot, fig.cap="Meta-analysis volcano plot highlighting top candidate oncogenes and tumor suppressors"}
# Use gene symbols as labels for the volcano plot
meta_res_plot <- meta_result
meta_res_plot@meta_table$gene_id <- meta_res_plot@meta_table$symbol

mx_volcano(meta_res_plot, padj_threshold = 0.05, lfc_threshold = 1.0, label_top = 8)
```

## 9.2 Multi-Cohort Forest Plots

Forest plots display the individual cohort effect sizes ($\log_2\text{FC}$) with 95% Confidence Intervals alongside the summary DerSimonian-Laird Random Effects pooled diamond:

### Top Conserved Oncogene (`MST1R` / RON)

```{r forest-up, fig.cap="Multi-cohort forest plot for oncogene MST1R"}
mx_forest("4486", de_results, meta_result)
```

### Top Conserved Tumor Suppressor (`FAM107A` / DRR1)

```{r forest-down, fig.cap="Multi-cohort forest plot for tumor suppressor FAM107A"}
mx_forest("11170", de_results, meta_result)
```

---

# 10. Interactive Explorer & Export

## Launch Interactive Shiny Dashboard

Launch the GUI locally to explore volcano plots, adjust FDR thresholds, and inspect forest plots interactively:

```{r run-app, eval=FALSE}
# Launch local Shiny dashboard
metaXpress::mx_run_app()
```

Or test the deployed cloud explorer at [https://hossainlab.shinyapps.io/metaXpress-demo/](https://hossainlab.shinyapps.io/metaXpress-demo/).

## Export Tabular Results

```{r export-csv}
out_file <- mx_export(meta_result, format = "csv", output_dir = tempdir(), prefix = "pan_cancer_meta")
cat("Exported results table to:", out_file, "\n")
```

---

# 11. Session Information

```{r session-info}
mx_session_info()$session_info
```
