---
title: "metaXpress Quickstart"
author: "Md. Jubayer Hossain"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{metaXpress Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  eval     = FALSE,  # All chunks are illustrative; run interactively
  purl     = FALSE   # Exclude from R CMD check vignette code extraction
)
```

# Overview

`metaXpress` provides an end-to-end pipeline for integrating multiple bulk
RNA-seq studies into a single meta-analysis result. This vignette walks
through a minimal example using the built-in simulated dataset.

```{r load-package}
library(metaXpress)
```

# Load Example Data

The package ships with a small simulated dataset: three studies with 500 genes
each and partial gene overlap.

```{r load-data, eval=FALSE}
data(metaXpress_example)
studies <- metaXpress_example
lapply(studies, show)
```

# Step 1 — QC Filtering

Studies are already QC-scored in the example data. In a real workflow, QC
scores are computed by `mx_fetch_geo()` automatically, or manually via
`mx_qc_study()`.

```{r filter-studies}
studies <- mx_filter_studies(studies, qc_threshold = 7)
message("Studies passing QC: ", length(studies))
```

# Step 2 — Align Genes

Find the common gene universe across all studies.

```{r align-genes}
studies <- mx_align_genes(studies)
```

# Step 3 — Per-Study Differential Expression

Run DESeq2 on each study independently.

```{r de-analysis}
studies <- mx_de_all(studies, method = "DESeq2", formula = ~ condition)
mx_de_summary(studies)
```

# Step 4 — Meta-Analysis

Combine per-study DE results using the random effects model.

```{r meta-analysis}
de_results <- lapply(studies, function(s) s@de_result)
meta_result <- mx_meta(de_results, method = "random_effects")
meta_result
```

# Step 5 — Visualise

```{r volcano, fig.width=7, fig.height=5}
mx_volcano(meta_result, label_top = 10)
```

```{r forest, fig.width=6, fig.height=4}
# Forest plot for the top gene
top_gene <- meta_result@meta_table$gene_id[
  which.min(meta_result@meta_table$meta_padj)]
mx_forest(top_gene, de_results, meta_result)
```

```{r heterogeneity, fig.width=6, fig.height=4}
mx_heterogeneity_plot(meta_result)
```

# Step 6 — Export Results

```{r export, eval=FALSE}
mx_export(meta_result, format = "csv", output_dir = tempdir())
```

# Session Info

```{r session-info, eval=FALSE}
mx_session_info()$session_info
```
