---
title: "metaXpress Quickstart"
author: "Md. Jubayer Hossain"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{metaXpress Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  eval     = FALSE,  # All chunks are illustrative; run interactively
  purl     = FALSE   # Exclude from R CMD check vignette code extraction
)
```

# Overview

`metaXpress` provides an end-to-end pipeline for integrating multiple bulk
RNA-seq studies into a unified meta-analysis. This guide walks through installation
from GitHub and a minimal workflow for ingesting public data from NCBI GEO,
performing quality control, differential expression analysis, meta-analysis,
and visualization.

# Installation from GitHub

`metaXpress` is currently available on GitHub and targets future submission to Bioconductor.

You can install the package directly from GitHub using `remotes` or `pak`:

```{r install-github, eval=FALSE}
# Install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

# Install metaXpress directly from GitHub
remotes::install_github("hossainlab/metaXpress")
```

Alternatively, using `pak`:

```{r install-pak, eval=FALSE}
# install.packages("pak")
pak::pkg_install("hossainlab/metaXpress")
```

Once installed, load the library:

```{r load-package, eval=FALSE}
library(metaXpress)
```

# Ingest Public Data from GEO

You can fetch public RNA-seq count matrices and curated metadata directly from
NCBI GEO using accession numbers:

```{r load-data, eval=FALSE}
# Fetch two bulk RNA-seq cohorts from GEO
studies <- mx_fetch_geo(c("GSE53697", "GSE95587"))
lapply(studies, show)
```

# Step 1 — QC Filtering

Each study is evaluated against a 10-point quality control standard
(replicate counts, depth, alignment rate, raw integer counts, etc.):

```{r filter-studies, eval=FALSE}
# Filter studies passing QC threshold (score >= 7)
studies <- mx_filter_studies(studies, qc_threshold = 7)
message("Studies passing QC: ", length(studies))
```

# Step 2 — Align Gene Spaces

Identify the common gene universe across all cohorts:

```{r align-genes, eval=FALSE}
studies <- mx_align_genes(studies)
```

# Step 3 — Per-Study Differential Expression

Run negative binomial GLMs on each study independently using DESeq2, edgeR, or limma-voom:

```{r de-analysis, eval=FALSE}
studies <- mx_de_all(studies, method = "DESeq2", formula = ~ condition)
mx_de_summary(studies)
```

# Step 4 — Meta-Analysis

Combine per-study effect sizes and standard errors using the DerSimonian-Laird Random Effects Model:

```{r meta-analysis, eval=FALSE}
de_results <- lapply(studies, function(s) s@de_result)
meta_result <- mx_meta(de_results, method = "random_effects")
meta_result
```

# Step 5 — Visualise

Generate publication-grade plots:

```{r volcano, fig.width=7, fig.height=5, eval=FALSE}
# Volcano plot of meta-analysis results
mx_volcano(meta_result, label_top = 10)
```

```{r forest, fig.width=6, fig.height=4, eval=FALSE}
# Multi-study forest plot for the top gene
top_gene <- meta_result@meta_table$gene_id[
  which.min(meta_result@meta_table$meta_padj)]
mx_forest(top_gene, de_results, meta_result)
```

```{r heterogeneity, fig.width=6, fig.height=4, eval=FALSE}
# Between-study heterogeneity distribution
mx_heterogeneity_plot(meta_result)
```

# Step 6 — Interactive Shiny Explorer

You can also explore your meta-analysis results interactively in your browser:

```{r shiny-app, eval=FALSE}
# Launch local Shiny dashboard
metaXpress::mx_run_app()
```

Or test the online demo at [https://hossainlab.shinyapps.io/metaXpress-demo/](https://hossainlab.shinyapps.io/metaXpress-demo/).

# Step 7 — Export Results

```{r export, eval=FALSE}
mx_export(meta_result, format = "csv", output_dir = tempdir())
```

# Session Info

```{r session-info, eval=FALSE}
mx_session_info()$session_info
```
